Effective Reionization Likelihood from Planck 2018 Data (?)

### 2021-03-26 paper/rei.tex
Updates from Overleaf

\newcommand{\wh}[1]{\textcolor{blue}{#1}}
\newcommand{\ch}[1]{\textcolor{red}{#1}}

@@

The paper is structured as follows. We first describe in \S~\ref{sec:background} the background on the reionization principal components and the kernel density estimate (KDE) technique used for building the effective likelihood. Then in \S~\ref{sec:results}, we describe the PC results using the Planck 2018 + \texttt{srollv2} likelihood, and compare with previous results in literature. Then in \S~\ref{sec:effective_likelihood}, we present the effective likelihood code, and demonstrate its fidelity with two examples, 1) the standard steplike model, and 2) a two-parameter model where a plateau of ionization at high-$z$ is added to the standard one-step model for illustration purposes. Finally, we summarize our results and conclude in \S~\ref{sec:conclusion}.

@@

\section{PC Technique}
\label{sec:background}
The principal component technique for constraining reionization using the large-angle $C_\ell^{EE}$ polarization spectrum was first introduced in Ref.~\cite{Hu:2003gh}. We now briefly summarize the PC technique in \S~\ref{sec:PC}, as well as the kernel density estimate technique for building the effective likelihood in \S~\ref{sec:KDE}. We refer the readers to Refs.~\cite{
%Hu:2003gh, %Mortonson:2008rx, 
Mortonson:2007hq,Heinrich:2016ojb} respectively for a more complete description of these topics.

@@

The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$. Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~(\ref{eq:xetommu}). Since $\mathbf{m}$ can take any continuous value, we approximate its effective likelihood with a kernel density estimate
\beq
{\cal L}_{\rm KDE}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N_{\rm sample}} w_i K_f(\mathbf{m}-\mathbf{m}_i),
\eeq
where the overall normalization is arbitrary, and where we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero and covariance $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller than 1.
For a Gaussian posterior, increasing the covariance by $1+f$ corresponds to increasing the standard deviation by approximately $1+f/2$. To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \approx 1.1 \times 10^{6}$ chain samples. 
% 1.06 cut w/ physicality prior
% 1.072 no cut
For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient.
%Note also that we employ PC chains without the physicality priors allowing the smoothing kernel to cross over these priors.

@@

For any set of model parameters ${\bf p}$ with their given prior probability distribution
$P({\bf p})$, their posterior distribution can be approximated as 
\begin{equation}
P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p).
\end{equation}
with PC\,=\,KDE or Gaussian.
This approach obviates the need to implement specific reionization models into CAMB and conduct separate COSMOMC sampling for each, thereby significantly reducing both human and computational effort.
In \S~\ref{sec:effective_likelihood} we test the effective KDE and Gaussian likelihoods against an exact implementation
for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.

@@

Although reionization PCs are mainly a tool for model testing using complete information from CMB polarization, as they do not reconstruct $x_e(z)$ itself, in particular the its rapidly  aspects, they do provide model-independent constraints on redshift integrated quantities such as  the cumulative Thomson optical depth
\begin{align}
\tau(z,z_{\rm max}) & = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)}.
\label{eq:cumtau}
\end{align}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter. 
Given the tight constraints on cosmological parameters in the $\Lambda$CDM model, this quantity is well approximated by
\begin{align}
\tau_{\rm PC}(z,z_{\rm max})& = \sum_{a=1}^5 m_a \tau_a(z,z_{\rm max}) + \tau_{\rm fid}(z,z_{\rm max}),
\label{eq:cumtauPC}
\end{align}
where $\tau_a$ and $\tau_{\rm fid}$ are defined by employing Eq.~({\ref{eq:mmutoxe}}) in
Eq.~(\ref{eq:cumtau}) using the fiducial model 
for $H(z)$ and number densities.
 When we omit the redshift arguments, we
implicitly mean the total range, e.g. 
 $
 \tau_{\rm PC}=\tau_{\rm PC}(0,z_{\rm max}).
 $
 \ch{CH: what about recaling for different cosmological model?}

@@

\begin{figure}
\includegraphics[width=0.48\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
\caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $m_1-m_2$ PC plane. We also 
the Planck 2018 constraints using two different low-$\ell$ $EE$ likelihoods: the \texttt{simall\_EE} likelihood (red) and the \texttt{srollv2} likelihood (blue). The \texttt{srollv2} likelihood used better foreground removal and is what we use as our fiducial results throughout this paper. 
\ch{keeping srollv2 in label only here}
} \ch{keep same color scheme as Fig. 2, use green maybe for simallEE}
\label{fig:plot_m1m2_2015_vs_2018}
\end{figure}

In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we also show the constraints assuming a Gaussian posterior in $m_a$ using the PC chain covariance and means listed in Table~\ref{tab:PC_stats}.  The visual agreement with the full posteriors is remarkably good, especially in the well constrained components.  This suggests that the simple Gaussian  likelihood of Eq.~(\ref{eq:gaussian}) may suffice for rough constraints on models.    However we shall see that a quantitative comparison shows that there is enough skewness in the posteriors that the KDE likelihood 
still performs noticeably better.  \wh{perhaps shift the tone of this statement since we seem to be concluding that Gaussian is good enough.}

@@

We now turn to constraining reionization models with PCs using the
effective likelihood approach described in~\refsec.  The benefit of this approach is that
complete constraints on any model of the ionization history  between $6 \leq z \leq \zmax$ can now be obtained without reanalyzing the Planck data. 
This effective likelihood code is available on GitHub at \url{} [give URL]. The initial release corresponds to our fiducial constraints derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{srollv2}$ likelihood combination with $\zmax = 30$.

To use this approach, one simply specifies the functional form for $x_e(z)$ in the code, which then is converted to the PC parameters $m_a$ for the effective likelihood evaluation assuming the fiducial $\Lambda$CDM cosmology.   
To obtain reionization model parameter posteriors, one samples this likelihood with any desired priors.   
We have also provided an MCMC sampler for models of high dimension (currently in CosmoMC). Note that the model parameters or priors must be arranged to explicitly satisfy 
full ionization below $z=6$. To avoid over-smoothing parameter posteriors while maintaining accuracy during the KDE operation, we suggest using the default value of $f = 0.14$. \wh{Is Gaussian part of this package or a separate package? either way should mention this too.}

Next we demonstrate the use and the successful recovery of parameter posteriors using the effective likelihood code by comparing them to an exact MCMC analysis for two examples: 1) The tanh model, which is the standard approach used in CAMB; 2) a two-step toy model which has one additional parameter modeling a high-$z$ ionization plateau before the transition to full ionization in the canonical tanh model.

@@

\ch{quote tau tanh KDE and gauss numbers with labels like $\tau_{tanh} = ... $(KDE).}
In \reffig{tanh} we show the posterior distribution of $\tau$ for the effective likelihood results, in which the only sampled parameter is $\tau$ (all other cosmological parameters are fixed at the Planck best-fit model), vs the full likelihood results in which the five other $\Lambda$CDM parameters were also varied using the same priors for the $\tanh$ model.
%including $\tau$. 
%Because the PC method cannot accurately represent models that are not fully ionized below $z = 6$, we apply a prior $\zre < 6.1$ to both chains given the transition width of about $0.015(1+z) \approx 0.1$ at $z = 6$. 
The posteriors agree well; the effective likelihood gives a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing. We expect the smoothing to be roughly at the $\sim$7\% level in the parameter posterior corresponding to a smoothing factor of $1+f = 1.14$ applied to the covariance of the PCs used for smoothing kernel during KDE. Indeed, the KDE result $\tau = 0.0588 \pm 0.0066$ has a negligibly lower mean ($0.05\sigma$) and a 6.5\% larger standard deviation (as expected) than the full likelihood result $\tau = 0.0591 \pm 0.0062$.

Instead of the effective likelihood, we can alternately use its Gaussian approximation shown in Eq.~(\ref{eq:gaussian}). The Gaussian likelihood is faster to evaluate as it does not require looping over the entire $\sim 10^6$ points in the PC chain. [add timing here]
%and amounts to one evaluation of the multi-variate Gaussian function per sample in the reionization model parameter space. 
We obtain for the tanh model $\tau = 0.0594 \pm 0.0063$. While it is a good approximation (the mean and standard deviation are similar to those from the full likelihood and the effective likelihood using KDE), we see in Fig.~\ref{fig:tanh} that the peak of the distribution for the Gaussian likelihood is offset from the other two. This is because the Gaussian likelihood is by definition symmetric in the PC space and so it cannot capture any skewness in the PC distribution. As a result the tanh $\tau$ constraints are mostly symmetric too, $\tau = 0.0594_{-0.0064}^{+0.0064}$ if we consider the 68\% lower and upper limits, whereas the full KDE result $\tau = 0.0588_{-0.0072}^{+0.0060}$ is able to capture the skewness in the distribution
\wh{seems like we should quote the direct results to make the claim that it is correctly capturing skewness - also,  is there a reason why we start with symmetric errors instead of just giving these as the main results (maybe also a displayed equation to highlight)? if it's to quote the effect of 7\% smoothing we could still do that on the average of + and - errors.}
$\Delta \tau \sim 0.0012$, a level consistent with shifts in the peaks seen in Fig.~\ref{fig:tanh}. \ch{Add $\tau(15,\zmax)$ 95\% upper limit for tanh direct MCMC and KDE.}

\ch{[double check all these numbers again (some chains are longer now)]}.

@@

We also show in Fig.~\ref{fig:plot_taugtz_PC_vs_tanh} the cumulative optical depth obtained from directly integrating the tanh model in the exact likelihood chains vs that from the 5-PC representation of the tanh models in the effective likelihood chains using Eq.~(\ref{eq:cumtauPC}).
\wh{
Notice that because of the truncation of the PC expansion at $a=5$, this representation shows an small oscillatory cumulative $\tau$ at $z \gtrsim 10$ whereas the true model has negligible ionization there.   This is because the best constrained modes only represent the smooth features in the ionization history and truncation is similar to a low pass filter in Fourier space.  Though these zero mean, small amplitude oscillations do not bias results when averaged over redshift, this effect should be born in mind when interpreting the model independent results below.  
}

\wh{
In the exact or in the effective likelihood approaches (ignoring these oscillations),
we can see that the  tanh model does not allow for high redshift ionization by construction.   As we shall next see in the next example the data themselves do allow for a small amount of high redshift ionization when the form of the model permits them.
%As we shall see in \S \ref{sec:modelindependent} this is also captured in our model independent approach to cumulative optical depth constraints. 
}

@@

\begin{figure}[t]
%\includegraphics[width=0.47\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
\includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
\caption{Comparison of the marginalized posteriors and the 2D 68\% and 95\% C.L. contours for the two-step model  described in \refsec{example2} between different likelihoods: The full Planck likelihoods (shaded), the effective likelihoods in the KDE mode with f = 0.14 (blue solid) and in the Gaussian approximation mode (black dashed). The results agree well with each other. As in Fig.~\ref{fig:tanh}, both the KDE and Gaussian modes agree very well with the full likelihood results, with the KDE  contours being slightly broader as expected. {To add in caption: best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (0.053, 0.006)$, as well as  a model at the 95\% C.L. $(\taulo, \tauhi) = (0.039, 0.020)$}  {Otherwise this plot is done.} \wh{Gaussian looks lumpy in the marginal distributions?  also in 2D Gaussian may arguably be better than KDE. \ch{double check they are all smoothed the same way, and number of chain points}}
}
\label{fig:two_parameter_model_2D}
\end{figure}

@@

In practice, we parameterize the model by $\taulo(\zre, \xemin)$ and $\tauhi(\zre, \xemin)$.  They are defined as in \refeq{cumtau} but with the boundaries $[0, z_{\rm split}]$ and $[\tau(z_{\rm split}, \infty]$ for $\taulo$ and $\tauhi$ respectively, where the split is chosen as $z_{\rm split} = \zre + \Delta z_2$ to be conservative on the preference of $\tauhi$ in the data. [double check this too.]
Given a model $(\taulo, \tauhi)$, we find the corresponding ($\zre, \xemin$) through an iterative search similar to that used in the canonical tanh model for finding $\zre$ given a total optical depth $\tau$. 
For both the full likelihood and effective likelihood runs, we adopt the flat priors on $\taulo, \tauhi, \tau_{\rm tot} \in [0, \tau_{\rm max}]$ where  $\tau_{\rm tot} = \taulo + \tauhi$ is the total optical depth, and $\tau_{\rm max} = 0.35$. 
%except that $\taulo \in [0.03, \tau_{\rm max}]$ for the effective likelihood run. [not sure if this last sentence is needed; since we put a zre prior anyways.]
We further apply a prior cut in $\zre$ space to keep models with $\zre > 6.1$ for both the full and effective likelihood chains as in the tanh model. Here we used again $f = 0.14$ in the KDE computation. \wh{if we never show anything but $f=0.14$ we could just say it's always this value in the method section}

In ~\reffig{two_parameter_model_2D}, we compare the 1D and 2D marginalized posteriors of $\taulo$ and $\tauhi$.\wh{The exact, KDE and Gaussian PC treatments again agree very well.} 
Note that in the canonical tanh model $\taulo \lesssim 0.04$ roughly corresponds to $\zre < 6$ so that the posteriors distort near this value.
The models in Fig.~\ref{fig:two_step_model} are represented by the star and cross symbols for the best fit case and a model allowed at the 95\% CL.

% Not sure we need to get into this subtlety - the caveat being that getdist will smooth the boundary even if it is in reality sharper.   Not sure we know which effect dominates and it probably doesn't matter much:
%
%; for the two-step model, the correspondence between $\xemin$ and $\zre$ is not one-to-one and depends on the value of $\tauhi$. This is why the cut on the 2D constraint ellipses at low $\taulo$ is of a curved shape rather than a straight cut, because the $\zre$ prior corresponds to a different value of $\taulo$ for a different $\tauhi$.

[add timing results here?]

[comment on effective likelihood w/ KDE vs Gaussian likelihood?]

The marginalized 1D constraints from the effective likelihood $\taulo = XX \pm XX$ and $\tauhi = XX \pm XX$ are consistent with those of the full likelihood $\taulo = XX \pm XX$ and $\tauhi = XX \pm XX$. 
The 95\% C.L. upper limit on high-redshift optical depth are also consistent: $\tau(15, 30) < XX$ for the effective likelihood and $\tau(15, 30) < XX$ for the full likelihood.
\wh{Note also that the model 95\% C.L. allowed model in 
Fig.~\ref{fig:two_step_model} has
$\tau(15,30)= ZZ$.}

>>>>

\newcommand{\wh}[1]{\textcolor{blue}{#1}}
\newcommand{\ch}[1]{\textcolor{red}{#1}}
\definecolor{darkgreen}{rgb}{0.0, 0.2, 0.13}
\newcommand{\todo}[1]{\textcolor{darkgreen}{\bf{#1}}}

@@

The paper is structured as follows. We first describe in \S\ref{sec:background} the background on the reionization principal components and the kernel density estimate (KDE) technique used for building the effective likelihood. Then in \S\ref{sec:results}, we describe the PC results using the Planck 2018 + \texttt{srollv2} likelihood, and compare with previous results in literature. Then in \S\ref{sec:effective_likelihood}, we present the effective likelihood code, and demonstrate its fidelity with two examples, 1) the standard steplike model, and 2) a two-parameter model where a plateau of ionization at high-$z$ is added to the standard one-step model for illustration purposes. Finally, we summarize our results and conclude in \S\ref{sec:conclusion}.

@@

\section{PC Technique}
\label{sec:background}
The principal component technique for constraining reionization using the large-angle $C_\ell^{EE}$ polarization spectrum was first introduced in Ref.~\cite{Hu:2003gh}. We now briefly summarize the PC technique in \S\ref{sec:PC}, as well as the kernel density estimate technique for building the effective likelihood in \S\ref{sec:KDE}. We refer the readers to Refs.~\cite{
%Hu:2003gh, %Mortonson:2008rx, 
Mortonson:2007hq,Heinrich:2016ojb} respectively for a more complete description of these topics.

@@

The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$. Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~(\ref{eq:xetommu}). Since $\mathbf{m}$ can take any continuous value, we approximate its effective likelihood with a kernel density estimate
\beq
{\cal L}_{\rm KDE}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N_{\rm sample}} w_i K_f(\mathbf{m}-\mathbf{m}_i),
\eeq
where the overall normalization is arbitrary, and where we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero and covariance $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller than 1.
For a Gaussian posterior, increasing the covariance by $1+f$ corresponds to increasing the standard deviation by approximately $1+f/2$. To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \approx 1.1 \times 10^{6}$ chain samples. 
\ch{For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient for most models, and indeed we have tested it to work well for the ones we study in \S~\ref{sec:effective_likelihood}.}

@@

For any set of model parameters ${\bf p}$ with their given prior probability distribution
$P({\bf p})$, their posterior distribution can be approximated as 
\begin{equation}
P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p).
\end{equation}
with PC\,=\,KDE or Gaussian.
This approach obviates the need to implement specific reionization models into CAMB and conduct separate COSMOMC sampling for each, thereby significantly reducing both human and computational effort.
In \S\ref{sec:effective_likelihood} we test the effective KDE and Gaussian likelihoods against an exact implementation
for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.

@@

Although reionization PCs are mainly a tool for model testing using complete information from CMB polarization, as they do not reconstruct $x_e(z)$ itself, in particular the its rapidly  aspects, they do provide model-independent constraints on redshift integrated quantities such as  the cumulative Thomson optical depth
\begin{align}
\tau(z,z_{\rm max}) & = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)}.
\label{eq:cumtau}
\end{align}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter. 
Given the tight constraints on cosmological parameters in the $\Lambda$CDM model, this quantity is well approximated by
\begin{align}
\tau_{\rm PC}(z,z_{\rm max})& = \sum_{a=1}^5 m_a \tau_a(z,z_{\rm max}) + \tau_{\rm fid}(z,z_{\rm max}),
\label{eq:cumtauPC}
\end{align}
where $\tau_a$ and $\tau_{\rm fid}$ are defined by employing Eq.~({\ref{eq:mmutoxe}}) in
Eq.~(\ref{eq:cumtau}) using the fiducial model 
for $H(z)$ and number densities.
 When we omit the redshift arguments, we
implicitly mean the total range, e.g. 
 $
 \tau_{\rm PC}=\tau_{\rm PC}(0,z_{\rm max}).
 $
 \ch{CH: what about recaling for different cosmological model? Our tau(z,zmax) plots have cosmological parameters marginalized, while Figs 9-10 (total tau posterio) do not.}
 \wh{They probably shouldn't, i.e should be constructed directly out of $m_a$, especially if the increased uncertainty is small.   If it is noticable we'll need to comment on this since our recommended approach is to just use $m_a$ (which has cosmology premarginalized)
 for other constraints and fix cosmology thereafter.}

@@

\begin{figure}
\includegraphics[width=0.48\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
\caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $m_1-m_2$ PC plane. We also 
the Planck 2018 constraints using two different low-$\ell$ $EE$ likelihoods: the \texttt{simall\_EE} likelihood (red) and the \texttt{srollv2} likelihood (blue). The \texttt{srollv2} likelihood used better foreground removal and is what we use as our fiducial results throughout this paper. 
} \ch{keep same color scheme as Fig. 2, use green maybe for simallEE}
\label{fig:plot_m1m2_2015_vs_2018}
\end{figure}

In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we also show the constraints assuming a Gaussian posterior in $m_a$ using the PC chain covariance and means listed in Table~\ref{tab:PC_stats}.  The visual agreement with the full posteriors is remarkably good, especially in the well-constrained components.  This suggests that the simple Gaussian likelihood of Eq.~(\ref{eq:gaussian}) may suffice for rough constraints on models. \ch{We shall see in \S\ref{sec:effective_likelihood} that a quantitative comparison indeed shows good agreement between the KDE and Gaussian likelihoods, with only minor differences.}

@@

We now turn to constraining reionization models with PCs using the
effective likelihood approach described in~\S\ref{sec:effective_likelihood}.  \ch{The benefit of this approach is that
complete constraints on any model of the ionization history  between $6 \leq z \leq \zmax$ are obtained ahead of time and compressed into a simpler likelihood, so that when doing model testing one no longer needs to modify the $x_e(z)$ function in for example CAMB, each time a new ionization model needs to be tested. These provide savings on human time in addition to the computing speed-ups from the likelihood evaluations.}

\ch{Our code is available on GitHub at \url{...} [give URL]. 
The initial release corresponds to the fiducial results in this paper, i.e. derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{srollv2}$ likelihoods at $\zmax = 30$. 
The most important part of the code release is the python package called \texttt{Relike} which implements the Gaussian likelihood. The user can specify a customized $x_e(z)$ function taking in reionization model parameters and use \textt{Relike} to return the effective likelihood of the model. The returned likelihood is already marginalized over cosmological parameters. 
To obtain marginalized distributions more easily for high-dimensional parameter space, one may choose to sample \texttt{Relike} within a MCMC sampler such as Cobaya or Cosmosis \ch{[cite]}}

\ch{For convenience, we also provide a pre-assembled CosmoMC-Relike code using the CosmoMC sampler in generic mode, with which we have generated the results presented in this section. This CosmoMC-Relike code has an internal  implementation of the Gaussian and KDE likelihoods in fortran, which is separate from our main python package that only supports the Gaussian likelihood. }

\ch{When using the KDE likelihood, we suggest using the default value of $f = 0.14$ to avoid over-smoothing parameter posteriors while maintaining accuracy during the KDE operation. All our KDE results in this section are computed with $f = 0.14$, which we shall soon show to work well in recovering results from an exact MCMC. Note that in both the Gaussian and the KDE modes, the model parameters or priors must be arranged to explicitly satisfy fully-ionized hydrogen and singly-ionized helium for $z\leq 6$. }

Next we demonstrate the use and the successful recovery of parameter posteriors using the effective likelihood code by comparing them to an exact MCMC analysis using the original Planck likelihoods for two examples: 1) The tanh model, which is the standard approach adopted in CAMB; 2) a two-step toy model which has one additional parameter allowing for a high-$z$ ionization plateau before the transition to full ionization in the canonical tanh model.

@@

In \reffig{tanh} we show the posterior distribution of $\tau$ for the effective likelihood results, in which the only sampled parameter is $\tau$ (all other cosmological parameters are fixed at the Planck best-fit model), vs the exact Planck likelihood results in which the five other $\Lambda$CDM parameters were also varied using their standard priors.
The posteriors agree well; the effective likelihood gives a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing. We expect the smoothing to be roughly at the $\sim$7\% level in the parameter posterior corresponding to the smoothing factor of $1+f = 1.14$ we applied to the PC covariance during KDE. Indeed, the KDE result  
%
\beq
\tau_{\rm tanh} = 0.0588_{-0.0066}^{+0.0061}\;\; (\mathrm{KDE}) 
% before extending chains
%\tau_{\rm tanh} = 0.0588_{-0.0072}^{+0.0061}\;\; (\mathrm{KDE}) 
\eeq
%
has a negligibly lower mean ($0.05\sigma$) and a 4\% larger 68\% confidence region  than the exact likelihood result 
\beq
\tau_{\mathrm{tanh}} = 0.0591_{-0.0068}^{+0.0054}\;\; (\mathrm{Exact}).
\eeq
}
%before
%$\tau = 0.0591 \pm 0.0062$ (exact)
%$\tau = 0.0588 \pm 0.0066$ (KDE)
%now (possibly due to longer chains)
%$\tau = 0.0591 \pm 0.0063$ (exact)
%$\tau = 0.0588 \pm 0.0064$ (KDE)

Instead of the effective likelihood, we can alternately use its Gaussian approximation shown in Eq.~(\ref{eq:gaussian}). The Gaussian likelihood is faster to evaluate as it does not require looping over the entire $\sim 10^6$ points in the PC chain. [add timing here]

\ch{
We obtain for the tanh model 
\beq
\tau_{\rm tanh} = 0.0592_{-0.0062}^{+0.0062}\;\; (\mathrm{Gaussian})
\eeq
}
\ch{CH: need to rewrite the following paragraph. Double check that results do not change varying length of chains ...}
 While it is a good approximation (the mean and standard deviation are similar to those from the full likelihood and the effective likelihood using KDE), we see in Fig.~\ref{fig:tanh} that the peak of the distribution for the Gaussian likelihood is offset from the other two. This is because the Gaussian likelihood is by definition symmetric in the PC space and so it cannot capture any skewness in the PC distribution. As a result the tanh $\tau$ constraints are mostly symmetric too, $\tau = 0.0594_{-0.0064}^{+0.0064}$ \wh{this differs from above}
 if we consider the 68\% lower and upper limits, whereas the full KDE result 
\beq
\tau_{\rm tanh} = 0.0588_{-0.0072}^{+0.0060}\;\; (\mathrm{KDE})
\eeq
is able to capture the skewness in the distribution
%\wh{seems like we should quote the direct results to make the claim that it is correctly capturing skewness - also,  is there a reason why we start with symmetric errors instead of just giving these as the main results (maybe also a displayed equation to highlight)? if it's to quote the effect of 7\% smoothing we could still do that on the average of + and - errors.}
$\Delta \tau \sim 0.0012$, a level consistent with shifts in the peaks seen in Fig.~\ref{fig:tanh}. \ch{CH: To turn into a sentence: $\tau_{\rm tanh}(15,\zmax) < 0.002$ for both effectivey likelihoods and $\tau_{\rm tanh}(15,\zmax) = 0$ from exact MCMC.}
% 95\% UL:
%tanh Gauss: 1.996574506651048464e-03
%tanh KDE: 1.996656635642191091e-03
%tanh exact: 0.000000000000000000e+00

@@

We also show in Fig.~\ref{fig:plot_taugtz_PC_vs_tanh} the cumulative optical depth obtained from directly integrating the tanh model in the exact likelihood chains vs that from the 5-PC representation of the tanh models in the effective likelihood chains using Eq.~(\ref{eq:cumtauPC}).
Notice that because of the truncation of the PC expansion at $a=5$, this representation shows an small oscillatory cumulative $\tau$ at $z \gtrsim 10$ whereas the true model has negligible ionization there.   This is because the best constrained modes only represent the smooth features in the ionization history and truncation is similar to a low pass filter in Fourier space.  Though these zero mean, small amplitude oscillations do not bias results when averaged over redshift, this effect should be born in mind when interpreting the model-independent results in \S\ref{sec:modelindependent}.  
In the exact or in the effective likelihood approaches (ignoring these oscillations,
we can see that the tanh model does not allow for high redshift ionization by construction.  As we shall next see in the next example the data themselves do allow for a small amount of high redshift ionization when the form of the model permits them.
}

@@

\begin{figure}[t]
%\includegraphics[width=0.47\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
\includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
\caption{Comparison of the marginalized posteriors and the 2D 68\% and 95\% C.L. contours for the two-step model  described in \refsec{example2} between different likelihoods: The full Planck likelihoods (shaded), the effective likelihoods in the KDE mode with f = 0.14 (blue solid) and in the Gaussian approximation mode (black dashed). The results agree well with each other. As in Fig.~\ref{fig:tanh}, both the KDE and Gaussian modes agree very well with the full likelihood results, with the KDE  contours being slightly broader as expected. {To add in caption: best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (0.053, 0.006)$, as well as  a model at the 95\% C.L. $(\taulo, \tauhi) = (0.039, 0.020)$}  {Otherwise this plot is done.} \wh{Gaussian looks lumpy in the marginal distributions?  also in 2D Gaussian may arguably be better than KDE. \ch{double checked that they are all smoothed with default getdist setting; many more chains for Gaussian, running KDE to longer chains to compare afterward.}}
}
\label{fig:two_parameter_model_2D}
\end{figure}

@@

In practice, we parameterize the model by $\taulo(\zre, \xemin)$ and $\tauhi(\zre, \xemin)$.  They are defined as in \refeq{cumtau} but with the boundaries $[0, z_{\rm split}]$ and $[\tau(z_{\rm split}, \infty]$ for $\taulo$ and $\tauhi$ respectively, where the split is chosen as $z_{\rm split} = \zre + \Delta z_2$ to be conservative on the preference of $\tauhi$ in the data. [double check this too.]
Given a model $(\taulo, \tauhi)$, we find the corresponding ($\zre, \xemin$) through an iterative search similar to that used in the canonical tanh model for finding $\zre$ given a total optical depth $\tau$. 
For both the full likelihood and effective likelihood runs, we adopt the flat priors on $\taulo, \tauhi, \tau_{\rm tot} \in [0, \tau_{\rm max}]$ where  $\tau_{\rm tot} = \taulo + \tauhi$ is the total optical depth, and $\tau_{\rm max} = 0.35$. 
%except that $\taulo \in [0.03, \tau_{\rm max}]$ for the effective likelihood run. [not sure if this last sentence is needed; since we put a zre prior anyways.]
We further apply a prior cut in $\zre$ space to keep models with $\zre > 6.1$ for both the full and effective likelihood chains as in the tanh model.

In ~\reffig{two_parameter_model_2D}, we compare the 1D and 2D marginalized posteriors of $\taulo$ and $\tauhi$. The exact, KDE and Gaussian PC treatments again agree very well. 
Note that in the canonical tanh model $\taulo \lesssim 0.04$ roughly corresponds to $\zre < 6$ so that the posteriors distort near this value.
The models in Fig.~\ref{fig:two_step_model} are represented by the star and cross symbols for the best fit case and a model allowed at the 95\% CL.

\todo{add timing results somewhere}

\ch{The marginalized 1D constraints from the KDE and Gaussian likelihoods 
\beq
\taulo = 0.0525^{+0.0062}_{-0.0079}, \; \tauhi < 0.021\; \mathrm{(95\%\; C.L.)\; (KDE)}
\eeq 
and
\beq
\taulo = 0.0528^{+0.0064}_{- 0.0077}, \; \tauhi < 0.021\; \mathrm{(95\%\; C.L.)\; (Gaussian)}
\eeq 
are consistent with those from the exact likelihood 
\beq
\taulo = 0.0526^{+0.0058}_{-0.0078}, \; \tauhi < 0.021\; \mathrm{(95\%\; C.L.)\; (Exact).}
\eeq 
}
%exact: tauhi < 0.0206
%KDE: tauhi < 0.0214
%Gaussian: tauhi < 0.0207

% running longer KDE chains (numbers might be updated)

\ch{The 95\% C.L. upper limit on high-redshift optical depth are also consistent: $\tau(15, 30) < 0.017$ for both effective likelihoods and $\tau(15, 30) < 0.016$ for the exact likelihood.}
\ch{Note also that the 95\% C.L. allowed two-step model in 
Fig.~\ref{fig:two_step_model} has
$\tau(15,30)= 0.013$.}
%0.01343375 evaluated at fiducial cosmology

The Planck 2018 paper~\cite{Aghanim:2018eyx} also constrained $\tau(15, 30)$ in a model-independent way using the FlexKnot method, in which $x_e(z_i)$ at the $i$-th knots are varied as free parameters as well as the number of total knots. They obtained $\tau(15, 30) < 0.007$ (95\% C.L.), which would strongly rule out models that are still allowed at 95\% C.L. in this class. 
 In principle, the FlexKnot method allows for all possible physical ionization histories that increase monotonically with redshifts, and would include the two-parameter model described here. But apparently the implicit prior employed by the method disfavors optical depth at high redshift more than a flat prior in 
$\tau_{\rm hi}$ does.  We shall see in \S \ref{sec:modelindependent} that our model independent PC approach models such a flat prior better and of course the effective likelihood approach allows for any desired prior.
